---
title: "00.data"
author: "Guillaume Papuga"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, include=FALSE, message=FALSE, warning=FALSE)
```

```{r packages}
require(ade4)
require(here)
library(tidyverse)
library(stringr)
```

# Introduction
## Article
Data processed in this file belongs to a project of ecological niche analysis of two sister species of the genus Linum. 

## Format
This document is used to format data. No analysis is coded here.
All data names once processed follow the form d.something (d stands for "data", and the second part must clearly refer to the type of data).

# Population data
```{r}
# load raw data
d.pop = read.csv(here::here("data","raw", "data.station.csv"), sep = ";", dec = ",")

# process & correct
nb.row = nrow(d.pop)

# write data in the "processed" folder
write.table (x = d.pop, here::here("data","processed","d.pop.txt"), sep = ";", dec = ".")
```

The original dataset contains `r nb.row` populations.

# Habitat data
```{r}
# load raw data
d.hab = read.csv(here::here("data","raw", "bhab.csv"), sep = ";", dec = ",")

# process & correct
nb.row = nrow(d.hab)

# write data in the "processed" folder
write.table (x = d.hab, here::here("data","processed","d.hab.txt"), sep = ";", dec = ".")
```

The *habitat* dataset contains `r nb.row` populations.

# Soil data
```{r}
# load raw data
d.soil = read.csv(here::here("data","raw", "soil.csv"), sep = ";", dec = ",")

# process & correct
nb.row = nrow(d.soil)

# write data in the "processed" folder
write.table (x = d.soil, here::here("data","processed","d.soil.txt"), sep = ";", dec = ".")
```

The *soil* dataset contains `r nb.row` populations.

# Flower data
```{r}
# load raw data
d.flo = read.csv(here::here("data","raw", "flowers.csv"), sep = ";", dec = ",")

# process & correct
nb.row = nrow(d.flo)

# write data in the "processed" folder
write.table (x = d.flo, here::here("data","processed","d.flo.txt"), sep = ";", dec = ".")
```

The *flower* dataset contains `r nb.row` populations. Number of short and long style flowers.

# Synthesis : field data
```{r}
# Who's missing? 
# Some populations haven't been sampled for some traits (e.g. soil), so we need to know which one

```


# Occurences (databases)
The aim of this section is to build a database of occurence throughout the two species range in order to compute their climatic niche based on WorldClim data. 

## 1. Dataset structure
```{r}
# The matrix is named `d.occ` and must contain 6 columns
# code.pop : population code (dataset specific)
# sp.name : species names (either linum tenuifolium or linum suffruticosum)
# date : the date (yyyymmdd)
# x : coordinate on the x-axis (longitude in decimal degree)
# y : coordinate on the y-axis (latitude in decimal degree)
# precision : precision of the location (expressed in meter)
# source : explicit name of the source of the imported dataset

d.occ = as.data.frame(matrix (ncol = 7,nrow = 0))
colnames (d.occ) = c("code.pop", "sp.name", "date", "x", "y", "precision", "source")
```

## 2. Import
To date, we have gathered one dataset.

### 2a. CBN data
```{r}
# load raw data
d.occ.cbn = read.csv(here::here("data","raw", "bdd.psce.csv"), sep = ";", dec = ",")

# select correct columns
d.occ.cbn = d.occ.cbn[,c("code", 
                         "nom_retenu", 
                         "date", 
                         "longitude", 
                         "latitude", 
                         "precision.m", 
                         "organisation")]
colnames(d.occ.cbn) = colnames (d.occ)

# process & correct
d.occ.cbn = unique (d.occ.cbn) # delete duplicates
## correct the date

## correct the name of the species
species.id = unique (d.occ.cbn$nom_retenu)

## count observation
nb.row = nrow(d.occ.cbn)
```

This dataset contains `r nb.row` populations.

### 2b. iNat
```{r}

```

### 2c. Coimbra data (Ana Afonso)

## 3. Synthesis
```{r}
# Copy and paste each dataset on the basic matrix
d.occ = rbind(d.occ, d.occ.cbn)

# Eliminate duplicates
d.occ = unique (d.occ)

# write data in the "processed" folder
write.table (x = d.occ, here::here("data","processed","d.occ.txt"), sep = ";", dec = ".")
```




